---
title: "Metabolic distribution: simulation"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Overview: Simulation model to assess how metabolic inequality influences ecosystem 
processes (i.e., respiration). We assume that microbial respiration increases 
nonlinearly with cellular enzyme levels according to a Michaelis–Menten function. 
Using this framework, we evaluate how distributional properties, not only variance 
but also skew, affect aggregate function by comparing a Gaussian distribution of 
enzyme levels to lognormal distributions with varying levels of skew (controlled 
by sigma, the standard deviation in log space). We further examine how these effects 
depend on the degree of nonlinearity in the response function, as determined by 
the half-saturation constant (Km).

## Clear workspace and set directory

```{r setup}
rm(list=ls())
getwd()
setwd("~/GitHub/MetabolicDistribution/")
```

## Load packages and functions

```{r setup}
library("png")
library("dplyr")
library("grid")
```

## Part 1. Set up model conditions for simulating enzymes and respiration

```{r}
# Set seed for reproducibility in all random number generations: rnorm() and rlnorm()
set.seed(42)

# Parameters for Michaelis-Menten dynamics and different distributions

n <- 10^6 # number of cells in sample

mean_norm <- 10^5 # mean enzyme level of Gaussian distribution: reference point 

sd_norm   <- 0.25 * mean_norm # sd of enzyme level for Gaussian distribution: reference point

Rmax <- 1 # maximum respiration rate for a cell

sigma_vals <- seq(0.01, 3.5, by = 0.1) # sd for enzyme level skewness (and variance) in lognormal space

Km_vals <- c(1e2, 1e3, 1e4, 1e5) # half-saturation constant (Km) values in units of enzyme level

f_mm <- function(x, Km, Rmax = 1) (Rmax * x) / (Km + x) # Michaelis-Menten function
```

## Part 2: Simulations of respiration with Gaussian and lognormal

```{r}
# Sample per-cell enzyme levels from a Gaussian distribution
activity_norm <- rnorm(n, mean = mean_norm, sd = sd_norm) # 10^6 enzyme observations

# Ensure all activity values are positive (required for Michaelis-Menten)
activity_norm_pos <- pmax(activity_norm, 10^-12) # 10^6 enzyme observations

# Apply Michaelis-Menten function to Gaussian-sampled enzyme levels
  # Computes average per-cell respiration rate for n cells at each Km
  # Total community respiration = [sum of individual rates] = [n * mean of individual rates]
  # Therefore, results scale directly to aggregate (summed) respiration for fixed n:
baseline_resp <- sapply(Km_vals, function(Km) mean(f_mm(activity_norm_pos, Km = Km, Rmax = Rmax)))

# Create empty matrices for output across Km and sigma combinations
  
  # Storage for the respiration bias values (I_R)
  # Each entry = lognormal / Gaussian
	resp_ratio_matrix <- matrix(NA, nrow = length(sigma_vals), ncol = length(Km_vals)) 

	# Storage for mean per-cell respiration under the lognormal distribution
  # These are the raw values before taking ratios
	Ef_lognorm_matrix <- matrix(NA, nrow = length(sigma_vals), ncol = length(Km_vals))

# Loop over Km and sigma to simulate respiration and calculate the index of respiration bias (I_R)
	
	for (k in seq_along(Km_vals)) {
    Km <- Km_vals[k]
    
    for (i in seq_along(sigma_vals)) {
      sigma <- sigma_vals[i]
 
# Adjust lognormal parameters so the arithmetic mean matches the Gaussian mean as sigma changes: 
    # meanlog (mu) = mean of log-transformed values
    # Choose mu (meanlog) so the lognormal has arithmetic mean = mean_norm (Gaussian mean)
    # Arithmetic mean for a lognormal: E[X] = exp(mu + sigma^2 / 2)
    # Set equal to target mean: mean_norm = exp(mu + sigma^2 / 2)
    # Take natural log: log(mean_norm) = mu + sigma^2 / 2
    # Solve for mu: mu = log(mean_norm) - sigma^2 / 2
    # This allows skew (and variance) to vary with sigma while holding the mean constant
    # Any differences in respiration due to shape (heterogeneity), not mean enzyme activity
    mu <- log(mean_norm) - 0.5 * sigma^2 

# Sample per-cell enzyme levels from a lognormal distribution        
    # Represents multiplicative (skewed) variability among cells
    # Skewness and variance increase with sigma while the arithmetic mean is held constant
    activity_lognorm <- rlnorm(n, meanlog = mu, sdlog = sigma) # 10^6 enzyme observations

    # Ensure strictly positive values for numerical consistency with Gaussian case
    # (lognormal draws are already positive)
    activity_lognorm <- pmax(activity_lognorm, 10^-12)
    
# Apply Michaelis-Menten function to lognormal-sampled cells
  # Computes mean per-cell respiration with lognormal and current Km
  # Total community respiration = sum of individual rates = n * mean
  # Therefore, results scale directly to aggregate (summed) respiration for fixed n
  # Returns one value per loop iteration for the current Km-sigma combination
    mean_resp_lognorm <- mean(f_mm(activity_lognorm, Km = Km, Rmax = Rmax))

# Calculate index of respiration bias (I_R)
  # Ratio of aggregated respiration under lognormal vs. Gaussian activity (for the same Km)
  # Values > 1 indicate higher predicted community respiration under the lognormal case
  # Values < 1 indicate lower predicted community respiration under the lognormal case
    resp_ratio_matrix[i, k] <- mean_resp_lognorm / baseline_resp[k]
    
# Store mean per-cell respiration (scales to aggregate respiration via n * mean)
# Useful for plotting absolute values or checking numerical stability    
    Ef_lognorm_matrix[i, k] <- mean_resp_lognorm
    # Ef_lognorm_matrix
  }
}
```

## Part 3: Make figure

```{r}
# Index of respiration bias (I_R) as function of skew (sigma)
# Different colored lines for different Km values

png(filename="~/GitHub/MetabolicDistribution/output/6.simulation.png", width = 1600, 
    height = 1200, res = 150, bg = "white")

# Plot margins
par(mar = c(7, 9, 4, 2) + 0.1, mgp = c(4, 1, 0), lwd = 1.6)
par(pty = "s") # square plot

# Define sigma limtt for plotting
cutoff <- 3.2
idx <- sigma_vals <= cutoff

# Define colors
colors <- c("red","darkorange","darkgreen","blue")[1:length(Km_vals)] # for plotting

# Main plot
plot(sigma_vals[idx], resp_ratio_matrix[idx, 1],
     type = "l", lwd = 2.2, col = colors[1],
     ylim = range(resp_ratio_matrix, na.rm = TRUE),
     xlab = "", ylab = "",
     xaxt = "n", yaxt = "n",
     xlim = c(0, 3.7),
     cex.lab = 1.8)

# Bottom axis
axis(1, at = c(0,1,2,3), labels = c("0.0","1.0","2.0","3.0"),
     cex.axis = 1.6, lwd = 1.8, lwd.ticks = 1.8)

# Left axis
y_at <- pretty(range(resp_ratio_matrix, na.rm = TRUE), n = 5)
axis(2, at = y_at, labels = format(y_at, trim = TRUE, digits = 3),
     cex.axis = 1.6, lwd = 1.8, lwd.ticks = 1.8, las = 1)

# Add ticks: top (x) and right (y) axes
axis(3, at = c(0, 1, 2, 3), labels = FALSE, tck = -0.02,
     lwd = 1.8, lwd.ticks = 1.8)
axis(4, at = y_at, labels = FALSE, tck = -0.02,
     lwd = 1.8, lwd.ticks = 1.8)

# Add axis labels
mtext(side = 1, text = expression("Skewness (" * sigma * ")"),
      line = 4.5, cex = 1.95)
#mtext(side = 2, text = "Respiration ratio", line = 4.5, cex = 1.95)
mtext(side = 2,
      text = expression("Respiration bias" ~ (italic(I)[R])),
      line = 4.5,
      cex = 1.95)

# Add colored text next to lines for Km values
for (k in seq_along(Km_vals)) {
  lines(sigma_vals[idx], resp_ratio_matrix[idx, k], col = colors[k], lwd = 2)
  y_end <- tail(resp_ratio_matrix[idx, k], 1)
  x_end <- tail(sigma_vals[idx], 1)
  exp_val <- as.integer(round(log10(Km_vals[k])))
  
  # Improved label format: K[m] = 10^x
  label_expr <- as.expression(bquote(K[m] == 10^.(exp_val)))
  text(x_end + 0.02, y_end, labels = label_expr,
       col = colors[k], cex = 1.25, pos = 4)
}

abline(h = 1, col = "gray50", lty = 2)
box(lwd = 1.8)

# Close plot device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("~/GitHub/MetabolicDistribution/output/6.simulation.png")

grid.raster(img)
```
